---
title: "Spatial Auditing"
format: html
editor: visual
embed-resources: true
---

Example using the Ward of Ardwick in the Local Authority (LA) of Manchester shown as the thin black line in the map below. Ward has been chosen as it is a lot smaller than LA, but large enough for a meaningful analysis. 

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

library(sf)
library(tidyverse)
library(tmap)

region = "Ardwick"

wards = st_read("../play_survey/DAT/WD_DEC_2025_UK_BFC_-2745845960906252656.gpkg", quiet = TRUE)

# buffer to use in metres
buf = 1000

geo_in = filter(wards,WD25NM == region)

geo_buf = geo_in |> 
  st_buffer(buf)

tmap_mode("view")

tm_shape(geo_in) +
  tm_polygons(fill_alpha = 0, lwd = 2)
  

```

In this example the assessment will be carried out at Lower Super Output Area (LSOA). This is the spatial level that the census is carried out at and it is also the highest resolution population data that is available open access from the government. [Indicies of Multiple Deprevation (IMD) statistics](https://communitiesopendata-communities.hub.arcgis.com/) are also released annually at this spatial resolution.

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

IMD_2025 = st_read("../IMD/dat/LSOA_IMD2025_WGS84_-4854136717238973930.gpkg",quiet = TRUE) |> 
  st_transform(27700)

```

The map below shows the LSOAs within Ardwick, coloured by the IMD rank in the country (out of `r NROW(IMD_2025)`, lower is more deprived). Data for the audit will be imported for a 1km radius from the border of each LSOA. This is shown as the thicker black line.

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

IMD_poly = IMD_2025[geo_in,]

imd_lowest = which(min(IMD_poly$IMDRank) == IMD_poly$IMDRank)

geo_buf = IMD_poly |> 
  st_union() |> 
  st_buffer(buf)

tm_shape(geo_buf)+
  tm_polygons(alpha = 0,lwd = 3)+
tm_shape(geo_in) +
  tm_polygons(fill_alpha = 0,lwd = 2)+
  tm_shape(IMD_poly)+
  tm_polygons(fill = "IMDRank", 
              fill.scale = tm_scale_continuous(),
              lwd = 0.5)

```

For this example the following data will be used as the measure of play spaces:

- [OS Greenspace data](https://osdatahub.os.uk/data/downloads/open/OpenGreenspace)  
- [Open Street Map (OSM) Places of Interest (POIS)](https://wiki.openstreetmap.org/wiki/Map_features#Leisure)  
- [OSM Landuse](https://wiki.openstreetmap.org/wiki/Map_features#Landuse)  
- [OSM Natural](https://wiki.openstreetmap.org/wiki/Map_features#Natural)  
- OSM protected areas  

OS data was downloaded from the portal and OSM data imported using the [osmdata R package](https://github.com/ropensci/osmdata) imported and trimed to the area.

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}


path = "DAT/greater-manchester.gpkg"

all_play = bind_rows(
  st_read(path, layer = "gis_osm_pois_a_free", quiet = TRUE) |> 
    filter(fclass %in% c("playground", "park", "pitch", "track", "sports_centre", "sports_hall",
                         "swimming_pool", "fitness_centre", "ice_rink", "theme_park", "zoo",
                         "dog_park", "picnic_site", "community_centre", "camp_site", "arts_centre")) |>
    select(fclass, name, geom),
  
  st_read(path, layer = "gis_osm_landuse_a_free", quiet = TRUE) |> 
    filter(fclass %in% c("park", "recreation_ground", "grass", "meadow", "forest")) |>
    select(fclass, name, geom),
  
  st_read(path, layer = "gis_osm_natural_a_free", quiet = TRUE) |> 
    filter(fclass == "beach") |>
    select(fclass, name, geom),
  
  st_read(path, layer = "gis_osm_protected_areas_a_free", quiet = TRUE) |> 
    filter(fclass == "nature_reserve") |>
    transmute(fclass, name = NA, geom)
)


play_in = all_play |> 
  st_transform(27700) |> 
  st_intersection(geo_buf)

os_greenspace = st_read("DAT/opgrsp_gb.gpkg",layer = "greenspace_site", quiet = TRUE) |> 
  select(fclass = function.,name = distinctive_name_1,geom = geometry)

green_in = os_greenspace[geo_buf,]

osm_os = rbind(green_in,play_in) |> 
  group_by(fclass) |>
  summarise(geom = st_union(geom))


  tm_shape(geo_buf)+
  tm_polygons(alpha = 0,lwd = 3)+
    tm_shape(osm_os) +
  tm_polygons(fill = "fclass",
              fill.scale = tm_scale_categorical())


```
Importing road collisions from the STATS19 national police database for the last 5 years (2020 to 2024) using the [STATS19 R package](https://github.com/ropensci/stats19).

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

library(stats19)

# import crashes and trim to temporal parameters and make spatial
crashes_gb <- get_stats19("5 years", type = "collision") |> 
  filter(collision_year >= 2020 & collision_year <= 2024) |> 
  format_sf()

crashes_in = crashes_gb[geo_buf,] |> 
  mutate(severity = ifelse(collision_severity == "Fatal",1.5,collision_adjusted_severity_serious))

tm_shape(geo_buf)+
  tm_polygons(alpha = 0,lwd = 3)+
tm_shape(crashes_in)  +
  tm_symbols(
    size = "severity",
    fill = "number_of_casualties",
    col = "black",
    size.scale = tm_scale_continuous(values.range = c(0.6,1.2)),
    fill.scale = tm_scale_intervals(n = 5, values = "tol.incandescent"))


```

## Scoring
For the area within each LSOA and 1km from the border, the total area of play spaces is calculated along with the total collisions.

An example below is shown for LSOA `r IMD_poly[imd_lowest,]$LSOA21NM`, which ranks in the `r ceiling(IMD_poly[imd_lowest,]$IMDRank/NROW(IMD_2025)*100)`% of most deprived areas in Great Britain.

The border of the LSOA is shown as the thick black line. The collisions and play facilities are trimmed to the 1km radius from the border of the LSOA and the total collisions and play area in km2 summed up.

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

LSOA = IMD_poly[imd_lowest,]

  
  LSOA_1km = LSOA |> 
    select(LSOA21CD,IMDDecil,SHAPE) |> 
    st_buffer(buf)
  
  LSOA_play = st_intersection(osm_os,LSOA_1km) |> 
    mutate(tot_area_m2 = as.numeric(st_area(geom)))
  
  LSOA_crashes = crashes_in[LSOA_1km,] |> 
    filter(collision_severity %in% c("Serious","Fatal"))

  LSOA_1km$play_area = sum(LSOA_play$tot_area_m2)/1000000  

  LSOA_1km$rd_cas = sum(as.numeric(LSOA_crashes$number_of_casualties))  

  tm_shape(LSOA_1km)+
  tm_polygons(alpha = 0,lwd = 1)+
tm_shape(LSOA)+
  tm_polygons(alpha = 0,lwd = 3)+
  tm_shape(LSOA_play) +
  tm_polygons(fill = "fclass",
              fill.scale = tm_scale_categorical())+
tm_shape(LSOA_crashes)  +
  tm_symbols(
    fill = "number_of_casualties",
    fill.scale = tm_scale_intervals(n = 5, values = "tol.incandescent"))

```
In total this LSOA has `r round(LSOA_1km$play_area,2)`km2 of "play" space and `r LSOA_1km$rd_cas` killed or seriously (KSI) casualties.

Applying this to all LSOAs and applying a scoring system to the values.

The scoring for each LSOA is based on the IMD scale of 1-10, so play area and crashes scored on a scale of 1-10 by taking the max value of each and splitting it into 10 even levels from 0, where 10 is the best (IMD, lowest deprivation, crashes, the least crashes and play the highest play space).

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

lsoa_list = list()
 for(l in IMD_poly$LSOA21CD){
   
   LSOA = IMD_poly[IMD_poly$LSOA21CD == l,]
   
   LSOA = select(LSOA, LSOA21CD,LSOA21NM,IMDDecil,IMDRank,SHAPE)

  LSOA_1km = LSOA |> 
    st_buffer(buf)
  
  LSOA_play = st_intersection(osm_os,LSOA_1km) |> 
    mutate(tot_area_m2 = as.numeric(st_area(geom)))
  
  LSOA_crashes = crashes_in[LSOA_1km,] |> 
    filter(collision_severity %in% c("Serious","Fatal"))

  LSOA$play_area = sum(LSOA_play$tot_area_m2)/1000000  

  LSOA$rd_cas = sum(as.numeric(LSOA_crashes$number_of_casualties))
  
  lsoa_list[[l]] = LSOA
  
 }


lsoa_total = do.call(rbind,lsoa_list)

play_max = max(lsoa_total$play_area)
cas_max = max(lsoa_total$rd_cas)

# play score breaks
ps_breaks = seq(0,play_max,length.out = 11)
bk_labels = as.character(seq(1,10))

# crash score breaks
cra_breaks = seq(0,cas_max,length.out = 11)

lsoa_scores = lsoa_total |> 
  mutate(play_score = as.numeric(as.character(cut(play_area,
                                                  breaks = ps_breaks, labels = bk_labels)))) |> 
  mutate(crash_score = as.numeric(as.character(cut(rd_cas,
                                                   breaks = cra_breaks, labels = rev(bk_labels))))) |> 
  mutate(total_score = IMDDecil + play_score + crash_score,
         crash_play_score = play_score + crash_score)
  
  
library(reactable)
library(htmlwidgets)

lsoa_table = lsoa_scores |> 
  st_set_geometry(NULL) |> 
  transmute(`LSOA name` = LSOA21NM,
            `IMD Rank` = IMDRank,
            `Play Area (km2)` = round(play_area,2),
            `Road casualties` = rd_cas,
            `IMD score` = IMDDecil,
            `Play score` = play_score,
            `Crash score` = crash_score,
            `Total score` = round(total_score)) |> 
  dplyr::arrange(desc(`Total score`)) |>
  reactable(
    theme = reactableTheme(
     style = list(fontFamily = "Geist, system-ui, sans-serif", fontSize = "13px"),
  headerStyle = list(fontWeight = 600),
  cellStyle = list(lineHeight = "1.5")
  ),
  defaultColDef = colDef(
    header = function(value) gsub(".", " ", value, fixed = TRUE),
    cell = function(value) format(value, nsmall = 1),
    align = "center",
    minWidth = 30,
    headerStyle = list(background = "grey")
  ),
  columns = list(
    name = colDef(minWidth = 80)  # overrides the default
  ),
  bordered = TRUE,
  highlight = TRUE,
  defaultPageSize = 30
)

htmlwidgets::saveWidget(lsoa_table, paste0("OUT/TAB/lsoa_table.html"), selfcontained = TRUE)

lsoa_table

```

For LSOA areas population data [is available for every age and sex](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates).

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

lsoa_pop = readRDS("OUT/DAT/pop_age_sex.RDS")

lsoa_score_pop = left_join(lsoa_scores,lsoa_pop, by = c("LSOA21CD" = "LSOA.2021.Code"))

backgroup_map = "CartoDB.Positron"

 tmap_mode("view")
  # initial geometry extent to start map off
  tm1 = tm_shape(geo_buf) +
    tm_polygons(fill_alpha = 0,
                col_alpha = 0,
                popup.vars = FALSE,
                group = "LAs", 
                group.control = "hide")
  
  geo_df = expand.grid(age = unique(lsoa_pop$age_band),sex = unique(lsoa_pop$sex)) |> 
    mutate(age_sex = paste(sex,age))
  geo_type = geo_df$age_sex
  # create function to loop through
  tm1 = reduce(geo_type, function(map_obj, ct) {
    
    demo_df = filter(geo_df,age_sex == ct)
    
    pop_df = filter(lsoa_score_pop,age_band == demo_df$age & sex == demo_df$sex)
    
    map_obj = map_obj +
      tm_shape(pop_df) +
      tm_polygons(fill = "pop",
                  fill.scale = tm_scale_intervals(n = 10),
                  #fill.legend = tm_legend(show = FALSE),
                  lwd = 0.5,
                  group = ct,
                  group.control = "radio")
    
    map_obj
    
    
  }, .init = tm1)
  
  map_title = "18 and under population of LSOA areas split by age and sex"
  
  tm1 <- tm1 + 
    tm_basemap(backgroup_map, group.control = "none")+ 
    tm_title(map_title)+
    # tm_add_legend(
    #   type = "polygons",
    #   labels = names(qpal),
    #   fill = unname(qpal)
    # )+
    tm_view(control.collapse = FALSE,overlay.groups = geo_type[1])

tm1

```


```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}


summary = lsoa_score_pop |>
  st_set_geometry(NULL) |> 
  group_by(age_band, sex) |>
  summarise(
    weighted_total = weighted.mean(total_score, pop, na.rm = TRUE),
    weighted_play = weighted.mean(as.numeric(play_score), pop, na.rm = TRUE),
    weighted_crash = weighted.mean(as.numeric(crash_score), pop, na.rm = TRUE),
    total_pop = sum(pop, na.rm = TRUE),
    n_regions = n()
  )

M14_18 = filter(summary,age_band == "14-18" & sex == "Male")

```

Weighted summary by age and sex for the entire ward of `r region`. In this area there are `r M14_18$total_pop` Boys age 14 to 18 and the weighted mean score is `r round(M14_18$weighted_total,1)`.

```{r, include=TRUE, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
summary_table = summary |> 
  transmute(`Age band` = age_band,
            `Sex` = sex,
            `Weighted total` = round(weighted_total,1),
            `Weighted play` = round(weighted_play,1),
            `Weighted crash` = round(weighted_crash,1),
            `Total population` = total_pop) |> 
  dplyr::arrange(desc(`Weighted total`)) |>
  reactable(
    theme = reactableTheme(
     style = list(fontFamily = "Geist, system-ui, sans-serif", fontSize = "13px"),
  headerStyle = list(fontWeight = 600),
  cellStyle = list(lineHeight = "1.5")
  ),
  defaultColDef = colDef(
    header = function(value) gsub(".", " ", value, fixed = TRUE),
    cell = function(value) format(value, nsmall = 1),
    align = "center",
    minWidth = 30,
    headerStyle = list(background = "grey")
  ),
  columns = list(
    name = colDef(minWidth = 80)  # overrides the default
  ),
  bordered = TRUE,
  highlight = TRUE,
  defaultPageSize = 30
)

htmlwidgets::saveWidget(summary_table, paste0("OUT/TAB/summary_table.html"), selfcontained = TRUE)

summary_table
    
```
    



